---
title: "TUMS - Results"
subtitle: "Figure 3A - MetaCyc Groups"
author: "Kimberley Parkin"
---

# load packages
```{r}
library(tidyverse)
library(phyloseq)
library(DESeq2)
library(dplyr)
library(ggrepel)
```

```{r}
packageVersion('phyloseq')
packageVersion('DESeq2')
```

# create directories
```{r}
functdata1 <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data/microba_batch1/functional_profiles'
functdata2 <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data/microba_batch2/functional_profiles'


plotdir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/figures'
resdir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/results'
datadir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data'
```

# load in phyloseq object
```{r}
pseq <- readRDS(file.path(datadir, 'phyloseq_objects/TUMS_phyloseq_counts_complete.rds'))

# remove sequencing replicate samples
pseq <- subset_samples(pseq, TUMS_ID != "NA")

# subset to follow-up samples
pseq <- subset_samples(pseq, age == "18")
```

# get metadata table
```{r}
metadata <- pseq %>%
  sample_data() %>%
  data.frame()
```

# load in functional data
```{r}
profiles1 <- read.table(file.path(functdata1, '/by_sample/MetaCyc_group.samples.tsv'), header = T, sep = "\t", quote = "")
profiles2 <- read.table(file.path(functdata2, '/by_sample/MetaCyc_group.samples.tsv.gz'), header = T, sep = "\t", quote = "")
```

# merge profile data from batch 1 and 2
```{r}
# repeat samples for batch effect analysis: "BBD7192", "BBD7271", "BBH0550"
# rename repeat samples for easy identification and to avoid duplication of rows during merge
profiles2 <- profiles2 %>%
  dplyr::rename(BBD7192_2 = BBD7192,
                BBD7271_2 = BBD7271,
                BBH0550_2 = BBH0550)
```

```{r}
profiles <- full_join(profiles1, profiles2)

# full_join introduces NA values, which need to be converted to 0
profiles[is.na(profiles)] <- 0
```

# set row names
```{r}
# assign VariableID as rownames
rownames(profiles) = profiles$VariableID

# remove VariableID column
profiles <- profiles[,-1]

# convert to matrix
profiles <- as.matrix(profiles)
```

# remove pathways with a row sum of 0
```{r}
# some pathways are not present in any sample, so we need to remove these

to.remove = which(rowSums(profiles)==0)

profiles = profiles[-to.remove,]
```

# subset to follow-up sample functional profiles
```{r}
# sub-set profiles data to only include 6-month samples
profiles <- profiles[, colnames(profiles) %in% metadata$sample_ID] 
```

# DESeq2 analysis

## change classes to data frame
```{r}
counts_data <- as.matrix(profiles)
colData <- as.matrix(metadata)

# convert counts data to integers 

# make sure the rownames in colData matches the column names in countData
all(colnames(counts_data) %in% rownames(colData)) # should be TRUE

# reorder the rows of colData so it matches the columns of counts_data
colData = colData[match(colnames(counts_data), rownames(colData)),]

# make sure they are in the same order
all(colnames(counts_data) == rownames(colData)) # should be TRUE
```

# construct DESeqDataSet object
```{r}
dds <- DESeqDataSetFromMatrix(countData = round(counts_data, 0), # round to the nearest whole number as DESeq cannot use decimals
                              colData = colData, 
                              design = ~ randomisation_group)
```

## run test
```{r}
ds.wald <- DESeq(dds, test = "Wald", fitType = "parametric")
```

# get DESeq2 results

```{r}
taxtab <- data.frame(profiles)
taxtab$pathway <- rownames(taxtab) 

# get deseq results
res <- results(ds.wald)
res$pathway <- rownames(res)

res2 <- data.frame(results(ds.wald, contrast = c("randomisation_group", "A", "B")))
res2$pathway <- rownames(res2)

# join with taxonomy table
res2 <- plyr::join(res2, taxtab, by = "pathway")

# get significant results only
sigtab <- filter(res2, padj < 0.05)
```

# Calculate confidence interval
```{r}
res2$log2FoldChange_lower <- res2$log2FoldChange - 1.96 * res2$lfcSE
res2$log2FoldChange_upper <- res2$log2FoldChange + 1.96 * res2$lfcSE
```

## volcano plot
```{r fig.width= 8, fig.height=8}
# Set the threshold for significance
threshold <- 0.05

# Create a new column for whether the point is above the threshold
res2$passed_threshold <- res2$padj < threshold

# Sort the data by p-value in ascending order
res2 <- res2[order(res2$padj), ]

# Create the volcano plot with labels for points above the threshold
volcano_plot <- ggplot(res2, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = passed_threshold), size = 2) +
  geom_text_repel(data = head(subset(res2, passed_threshold), n = 10),
                  aes(label = pathway), hjust = 1.2, vjust = 1.2, size = 8) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +
  theme_bw(base_size = 20) +
  labs(x = "log2 fold change", y = "-log10(adjusted p-value)", color = "Above Threshold") +
  theme(legend.position = "bottom")

# Display the plot
print(volcano_plot)
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig3A.metacyc_groups.pdf'), device = "pdf", dpi = 300, width = 8, height = 8)
```

# session info
```{r}
sessionInfo()
```
