---
title: "TUMS - Results"
subtitle: "Figure 1 - Effectiveness of Intervention"
author: "Kimberley Parkin"
---

# load packages
```{r}
library(tidyverse)
library(ggstatsplot)
```

# create directories
```{r}
plotdir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/figures'
resdir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/results'
datadir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data'
```

# load in data
```{r}
# water quality data
water <- readRDS(file.path(datadir, '/water_quality/cleaned_data.rds'))
```

# re-code visit column as factor
```{r}
water <- water %>% 
  dplyr::mutate(visit = case_when(visit == 1 ~ "Pre-intervention",
                                  visit == 2 ~ "Post-intervention",
                                  T ~ NA_character_),
                visit = as.factor(visit),
                visit = relevel(visit, "Pre-intervention"))
```

# compare total chlorine levels between intervention groups
```{r}
g1 <- water %>%
  filter(randomisation_group == "A") %>%
  ggwithinstats(
    x = visit,
    y = Chlor_T,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(0.2, 1.4, 0.2)) +
  labs(y = "Total chlorine (mg/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14),
    legend.position = "none") +
   ggtitle("Intervention group")

g2 <- water %>%
  filter(randomisation_group == "B") %>%
  ggwithinstats(
    x = visit,
    y = Chlor_T,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(0.2, 1.4, 0.2)) +
  labs(y = "Total chlorine (mg/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none") +
   ggtitle("Control group")

combine_plots(
  plotlist = list(g1, g2))
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig1B.intervention_Cl.png'), device = "png", dpi = 300, width = 8, height = 5)
```

# compare disinfection by-products between intervention groups (CHBr3)
```{r}
g1 <- water %>%
  filter(randomisation_group == "A") %>%
  ggwithinstats(
    x = visit,
    y = CHBr3,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) + 
  labs(y = "CHBr3 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14),
    legend.position = "none") +
  ggtitle("Intervention group")

g2 <- water %>%
  filter(randomisation_group == "B") %>%
  ggwithinstats(
    x = visit,
    y = CHBr3,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) +
  labs(y = "CHBr3 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12),  # Set the font size for axis text
    axis.title = element_text(size = 14),
    legend.position = "none") +
   ggtitle("Control group")

combine_plots(
  plotlist = list(g1, g2))
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig1C.intervention_CHBr3.png'), device = "png", dpi = 300, width = 8, height = 5)
```

# compare disinfection by-products between intervention groups (CHBrCl2)
```{r}
g1 <- water %>%
  filter(randomisation_group == "A") %>%
  ggwithinstats(
    x = visit,
    y = CHBrCl2,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) + 
  labs(y = "CHBrCl2 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14),
    legend.position = "none") +
  ggtitle("Intervention group")

g2 <- water %>%
  filter(randomisation_group == "B") %>%
  ggwithinstats(
    x = visit,
    y = CHBrCl2,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) +
  labs(y = "CHBrCl2 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12),  # Set the font size for axis text
    axis.title = element_text(size = 14),
    legend.position = "none") +
   ggtitle("Control group")

combine_plots(
  plotlist = list(g1, g2))
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig1D.intervention_CHBrCl2.png'), device = "png", dpi = 300, width = 8, height = 5)
```

# compare disinfection by-products between intervention groups (CHClBr2)
```{r}
g1 <- water %>%
  filter(randomisation_group == "A") %>%
  ggwithinstats(
    x = visit,
    y = CHClBr2,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) + 
  labs(y = "CHClBr2 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14),
    legend.position = "none") +
  ggtitle("Intervention group")

g2 <- water %>%
  filter(randomisation_group == "B") %>%
  ggwithinstats(
    x = visit,
    y = CHClBr2,
    type = "p",
    results.subtitle = FALSE,
    point.path.args = list(alpha = 0.3, linetype = "solid"),
    violin.args = list(width = 0.5, alpha = 0.2),
    package = "ggsci",
    palette = "nrc_npg"
  ) +
  geom_violin(aes(group = visit, fill = visit), alpha = 0.2, width = 0.5) +
  scale_y_continuous(breaks = seq(10, 80, 10)) +
  coord_cartesian(ylim = c(0, 85)) +
  labs(y = "CHClBr2 (ug/L)") +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text = element_text(size = 12),  # Set the font size for axis text
    axis.title = element_text(size = 14),
    legend.position = "none") +
   ggtitle("Control group")

combine_plots(
  plotlist = list(g1, g2))
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig1E.intervention_CHClBr2.png'), device = "png", dpi = 300, width = 8, height = 5)
```

# session info
```{r}
sessionInfo()
```