---
title: "TUMS - Results"
subtitle: "Figure 2B and Supplementary Figure 7 - Beta Diversity"
author: "Kimberley Parkin"
---

# load packages
```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
```

# create directories
```{r}
plotdir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/figures'
resdir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/results'
datadir <- 'w://Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data'
```

# load in phyloseq object
```{r}
pseq <- readRDS(file.path(datadir, '/phyloseq_objects/TUMS_phyloseq_counts_complete.rds'))
```

```{r}
# remove sequencing replicate samples
pseq <- subset_samples(pseq, TUMS_ID != "NA")

# change 'age' variable to factor
sample_data(pseq)$age <- as.character(sample_data(pseq)$age)
```

# recode age and randomisation_group variables
```{r}
# get metadata from phyloseq object
metadata <- pseq %>%
  sample_data() %>%
  data.frame()
```

```{r}
metadata <- metadata %>% 
  dplyr::mutate(age = case_when(age == "6" ~ "6 months",
                                age == "18" ~ "18 months",
                                T ~ NA_character_),
                age = as.factor(age),
                age = relevel(age, "6 months"))

metadata <- metadata %>% 
  dplyr::mutate(randomisation_group = case_when(randomisation_group == "A" ~ "Treatment",
                                                randomisation_group == "B" ~ "Control",
                                                T ~ NA_character_),
                randomisation_group = as.factor(randomisation_group),
                randomisation_group = relevel(randomisation_group, "Control"))
```

# add metadata back into phyloseq object
```{r}
sample_data(pseq) <- metadata
```

*overall cohort*

# calculate beta diversity of overall cohort
```{r}
# ordination
pseq.ord <- ordinate(pseq, "NMDS", "bray")
```

# plot
```{r}
p <- plot_ordination(pseq, pseq.ord, color = "age", shape = "randomisation_group") +
  geom_point(size = 3) +
  labs(title = "Overall (n = 300)") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        text = element_text(size = 14),
        legend.text = element_text(size = 10),
        legend.title = element_blank())

p <- p + scale_shape_manual(values = c(17, 1))

p$layers <- p$layers[-1]

p
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'fig2B.betadiversity_overall.pdf'), device = "pdf", dpi = 300, width = 5, height = 5)
```

# PERMANOVA
# get adonis data
```{r}
meta <- pseq %>%
  sample_data() %>%
  data.frame()

otu.tab <- pseq %>%
  otu_table() %>%
  t() %>%
  data.frame()
```

# run adonis test
```{r}
ad.test <- adonis2(otu.tab ~ age,
                  data = meta,
                  method = "bray")
ad.test
```

```{r}
R2 <- ad.test$R2[1]
R2 # 11.3% of the variance in our data is explained by age, which significantly contributes to beta diversity
```

*6 month samples*

# calculate beta diversity of 6 month samples
```{r}
# subset to 6 month samples
pseq6 <- subset_samples(pseq, age == "6 months")
```

```{r}
# ordination
pseq.ord <- ordinate(pseq6, "NMDS", "bray")
```

# plot
```{r}
p <- plot_ordination(pseq6, pseq.ord, shape = "randomisation_group") +
  geom_point(size = 3, color = "#F8766D") +
  labs(title = "6 months (n = 170)") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        text = element_text(size = 14),
        legend.text = element_text(size = 10),
        legend.title = element_blank())

p <- p + scale_shape_manual(values = c(17, 1))

p$layers <- p$layers[-1]

p
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'supplement7A.betadiversity_6m.png'), device = "png", dpi = 300, width = 5, height = 5)
```

# PERMANOVA

# get adonis data
```{r}
meta <- pseq6 %>%
  sample_data() %>%
  data.frame()

otu.tab <- pseq6 %>%
  otu_table() %>%
  t() %>%
  data.frame()
```

# run adonis test
```{r}
ad.test <- adonis2(otu.tab ~ randomisation_group,
                  data = meta,
                  method = "bray")
ad.test
```

```{r}
R2 <- ad.test$R2[1]
R2 # 0.6% of the variance in our data is explained by treatment group, which does not significantly contributes to beta diversity
```

*18 month samples*

# calculate beta diversity of 18 month samples
```{r}
#subset to 6 month samples
pseq18 <- subset_samples(pseq, age == "18 months")
```

```{r}
# ordination
pseq.ord <- ordinate(pseq18, "NMDS", "bray")
```

# plot
```{r}
p <- plot_ordination(pseq18, pseq.ord, shape = "randomisation_group") +
  geom_point(size = 3, color = "#00BFC4") +
  labs(title = "18 months (n = 130)") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        text = element_text(size = 14),
        legend.text = element_text(size = 10),
        legend.title = element_blank())

p <- p + scale_shape_manual(values = c(17, 1))

p$layers <- p$layers[-1]

p
```

```{r}
# save plot
#ggsave(file.path(plotdir, 'supplement7B.betadiversity_18m.png'), device = "png", dpi = 300, width = 5, height = 5)
```

# PERMANOVA

# get adonis data
```{r}
meta <- pseq18 %>%
  sample_data() %>%
  data.frame()

otu.tab <- pseq18 %>%
  otu_table() %>%
  t() %>%
  data.frame()
```

# run adonis test
```{r}
ad.test <- adonis2(otu.tab ~ randomisation_group,
                  data = meta,
                  method = "bray")
ad.test
```

```{r}
R2 <- ad.test$R2[1]
R2 # 0.6% of the variance in our data is explained by treatment group, which does not significantly contributes to beta diversity
```

# calculate confidence interval
```{r}
set.seed(123) # For reproducibility
n_boot <- 1000 # Number of bootstrap samples

# Create a function to calculate R2 for each bootstrap sample
bootstrap_R2 <- function(data, formula, dist_matrix, ...) {
  boot_data <- data[sample(nrow(data), replace = TRUE), ]
  adonis_boot <- adonis2(as.formula(formula), data = boot_data, permutations = 999, method = dist_matrix, ...)
  return(adonis_boot$R2[1])
}

# Running bootstrapping
boot_R2_values <- replicate(n_boot, 
                            bootstrap_R2(meta, "otu.tab ~ randomisation_group", "bray"))

# Calculate 95% confidence interval
ci_lower <- quantile(boot_R2_values, 0.025)
ci_upper <- quantile(boot_R2_values, 0.975)
ci_lower
ci_upper
```

# session info
```{r}
sessionInfo()
```