---
title: "TUMS - Supplementary Data"
subtitle: "Supplementary Figure 8 - Species Differential Abundance"
author: "Kimberley Parkin"
---

# In this script, we use the DESeq2 package to identify differentially abundant ASVs between groups.

# Set up
## Load packages
```{r}
library(phyloseq)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(DESeq2)
library(tidyverse)

# check package version
packageVersion("DESeq2")
```

## Load in functions
```{r}
# credit: Nelly Amenyogbe 
source('/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/R_functions/remove.low.counts.R')
```

## Load in phyloseq object as count data
```{r}
datadir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data'

ps <- readRDS(file.path(datadir, '/phyloseq_objects/TUMS_phyloseq_counts_complete.rds'))
```

## Get metadata from phyloseq object
```{r}
# Get metadata
meta <- ps %>%
  sample_data() %>%
  data.frame()
```

## Subset samples to remove NA values
```{r}
ps = subset_samples(ps, TUMS_ID != "NA")
ps = subset_samples(ps, age == "18")

table(sample_data(ps)$randomisation_group)
# There are 148 in the "A" group and 152 in the "B" group
```

# Filter OTU table to only include abundant ASVs
## Create matrix of ASVs from phyloseq object
```{r}
# Get OTU table
# we need our otu table transposed in order to use the remove_low_counts function
tab <- t(data.frame(otu_table(ps))) 
```

## Retain only ASVs present in at least 5% of samples
```{r}
# remove low counts
otu.f <- remove.low.counts(as.data.frame(tab), 1, 5)
  otu.f <- otu.f$df
  otu.f
```

## Filter phyloseq object to include relevant ASVs only
```{r}
ps.f <- subset_taxa(ps, rownames(tax_table(ps)) %in% colnames(otu.f))
```

# Run DESeq2
## Create DESeq2 objects
```{r}
dds <- phyloseq_to_deseq2(ps.f, design = ~ randomisation_group)

# Because there is a zero in almost every row, need to manually specify geoMeans  
# Source: https://support.bioconductor.org/p/62246/#62250

cts <- counts(dds)
geoMeans <- apply(cts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
dds <- estimateSizeFactors(dds, geoMeans = geoMeans)

dds
```

## Run test
```{r}
ds.wald <- DESeq(dds, test = "Wald", fitType = "parametric")
```

# Get deseq results
```{r}
taxtab <- data.frame(tax_table(ps))
taxtab$OTU.name <- rownames(taxtab) 

# Get deseq results
res <- data.frame(results(ds.wald))
res$OTU.name <- rownames(res)

# join with taxonomy table
res <- join(res, taxtab, by = "OTU.name")

# get significant results only
sigtab <- filter(res, padj < 0.01)
```

```{r}
dim(sigtab)
unique(sigtab$Genus)
# There are 55 differentially abundant genera between randomisation groups
```

# Plot results
## Arrange data
```{r}
# we will set our factor levels from the most to least affected genus (based log2FC values), and the phyla according to most to least represented
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels = names(x))
```


## Plot
```{r}
p <- ggplot(sigtab, aes(x = Genus, y = log2FoldChange, fill = Phylum)) +
  geom_point(size = 2, alpha = 0.8, color = "black", shape = 21) +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(face = "bold"),
        axis.line = element_line(size = 0.8),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        plot.margin = margin(t = 1, r = 5, b = 5, l = 5)) +  # Adjusting plot margins) +
  ggtitle("Genus") +
  labs(x = "", y = "Log2 Fold-Change") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#08519c")

p

#ggsave(file.path(plotdir, 'suppfig8_diff.abund.pdf'), device = "pdf", dpi = 300, width = 8, height = 8)
```

# Plot counts 1
```{r}
# visualize the top hits

# get a list of the most significant features
sigtab <- sigtab[order(sigtab$pvalue),]

features <- sigtab$OTU.name

plotCounts(dds = dds,
           gene = features[1],
           intgroup = "randomisation_group",
           transform = TRUE,
           normalized = TRUE)
```
# Plot counts 2
```{r}
# visualize the top hits

# get a list of the most significant features
sigtab <- sigtab[order(sigtab$pvalue),]

features <- sigtab$OTU.name

plotCounts(dds = dds,
           gene = features[2],
           intgroup = "randomisation_group",
           transform = TRUE,
           normalized = TRUE)
```

# Plot counts 3
```{r}
# visualize the top hits

# get a list of the most significant features
sigtab <- sigtab[order(sigtab$pvalue),]

features <- sigtab$OTU.name

plotCounts(dds = dds,
           gene = features[3],
           intgroup = "randomisation_group",
           transform = TRUE,
           normalized = TRUE)
```

# Plot counts 4
```{r}
# visualize the top hits

# get a list of the most significant features
sigtab <- sigtab[order(sigtab$pvalue),]

features <- sigtab$OTU.name

plotCounts(dds = dds,
           gene = features[4],
           intgroup = "randomisation_group",
           transform = TRUE,
           normalized = TRUE)
```


# Session info
```{r}
sessionInfo()
```

##### END