---
title: "TUMS - Supplementary Data"
subtitle: "Supplementary Figure 5 - Batch Effect"
author: "Kimberley Parkin"
---


*Description*

- Our TUMS samples were sequenced in two separate batches:
      - all the 6-month samples and some 18-month samples are in sequencing batch 1, 
      - and the remaining 18-month samples are in sequencing batch 2

- We randomly selected three 18-month sample libraries from batch 1 to be re-sequenced in batch 2 to check for batch effects


# Load packages 
```{r}
library(tidyverse)
library(Hmisc)
library(ggplot2)
library(ggfortify)
library(phyloseq)
```

# Create directories
```{r}
plotdir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/figures'
resdir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/src/TUMS/Paper_2_RCT/results'
datadir <- '/home/kparkin/file_workspace/Collaborative Projects/HPC_CEPI/ANALYSIS/Project_TUMS/data'

# sequencing batch 1: 170 6-month samples and 35 18-month samples (195 total)
# sequencing batch 2: 95 18-month samples (3 samples repeated from batch 1)

# total sequenced samples: 290 (287 unique samples)
# repeat samples for batch effect analysis: "BBD7192", "BBD7271", "BBH0550"
```

# Load in data
```{r}
# batch 1
qc1 <- read.table(file.path(datadir, '/microba_batch1/sample_qc.tsv'), header = T, sep = "\t", quote = "") 

# batch 2
qc2 <- read.table(file.path(datadir, '/microba_batch2/sample_qc.tsv'), header = T, sep = "\t", quote = "") 
```

```{r}
# add column to identify batch number
qc1 <- qc1 %>%
  add_column(batch = "1",
             .after = "ID")

qc2 <- qc2 %>%
  add_column(batch = "2",
             .after = "ID")
```

# Merge sequencing batches
```{r}
qc <- full_join(qc1, qc2)

# save qc data
saveRDS(qc, file.path(resdir, 'complete_qc.rds'))
```

# Add in metadata information
```{r}
# load in sample information table
meta_samples <- read.csv(file.path(datadir, '/metadata_tables/meta_samples.csv'), header = T)

# select age column from meta_samples data frame
sample_age <- select(meta_samples, c(sample_ID, age))

# rename ID column so it is the same as qc dataframe
sample_age <- dplyr::rename(sample_age, ID = sample_ID)

# add in age column to qc data frame
qc <- merge(sample_age, qc)
```

# Subset to only 18-month samples
```{r}
# subset only 18-month samples
qc <- subset(qc, age == "18")
```


# Plots

## Histograms
```{r}
# total mapped reads
p <- ggplot(qc, aes(x = STRINGENT_READS_MAPPED, fill = batch)) +
  geom_histogram(aes(y = after_stat(density)), 
                 colour = 1,
                 bins = 50) +
  facet_grid(. ~ batch) +                         
  geom_density(linewidth = 1.2, colour = "black", fill = NA)

print(p)

# save plot
ggsave(file.path(plotdir, 'qc_stringent_reads.pdf'), device = "pdf", dpi = 300, width = 8, height = 5)
```

```{r}
# total human DNA
p <- ggplot(qc, aes(x = TOTAL_HUMAN, fill = batch)) +
  geom_histogram(bins = 75) +
  facet_grid(. ~ batch) 

print(p)

# save plot
ggsave(file.path(plotdir, 'qc_human_dna.pdf'), device = "pdf", dpi = 300, width = 8, height = 5)
```

```{r}
# total assigned reads
p <- ggplot(qc, aes(x = MGDB_ASSIGNED_PERCENT, fill = batch)) +
  geom_histogram(aes(y = after_stat(density)), 
                 colour = 1,
                 bins = 50) +
  facet_grid(. ~ batch) +
  geom_density(linewidth = 1.2, colour = "black", fill = NA)

print(p)

# save plot
ggsave(file.path(plotdir, 'qc_mgdb_percent.pdf'), device = "pdf", dpi = 300, width = 8, height = 5)
```

# Check for batch effects

## Load in phyloseq object
```{r}
# load in phyloseq object
pseq <- readRDS(file.path(datadir, '/phyloseq_objects/TUMS_phyloseq_counts_complete.rds'))

# subset to 18-month samples only
pseq <- subset_samples(pseq, age == '18')
```

## Convert batch to character instead of numeric
```{r}
# access the sample data
sample_data <- pseq@sam_data

# convert the 'batch' column to character
sample_data$batch <- as.character(sample_data$batch)

# add batch number to sequencing replicates
sample_data$batch[is.na(sample_data$batch) & grepl("_2", rownames(sample_data))] <- "2"

# update the sample data in the phyloseq object
pseq@sam_data <- sample_data 
```

## Ordination
```{r}
pseq.ord <- ordinate(pseq, "NMDS", "bray")
```

## Plot ordination
```{r}
plot_ordination(pseq, pseq.ord, color = "batch") +
  theme_bw()

# save plot
ggsave(file.path(plotdir, 'qc_beta_diversity_batch.pdf'), device = "pdf", dpi = 300)
```

```{r}
p <- plot_ordination(pseq, 
                pseq.ord, 
                color = "sequencing_replicate") +
  theme_bw()

p

# save plot
ggsave(file.path(plotdir, 'qc_beta_diversity_replicate.pdf'), device = "pdf", dpi = 300)
```

# Taxa bar plots

```{r}
pseq_rep1 <- subset_samples(pseq, sequencing_replicate == '1')
p1 <- plot_bar(pseq_rep1, fill = "Phylum")

print(p1)

# save plot
ggsave(file.path(plotdir, 'qc_taxa_bar_plot_replicate1.pdf'), device = "pdf", dpi = 300)
```

```{r}
pseq_rep2 <- subset_samples(pseq, sequencing_replicate == '2')
p2 <- plot_bar(pseq_rep2, fill = "Phylum")

print(p2)

# save plot
ggsave(file.path(plotdir, 'qc_taxa_bar_plot_replicate2.pdf'), device = "pdf", dpi = 300)
```

# session info
```{r}
sessionInfo()
```